# Reproducible Research: Peer Assessment 1
Submitted by Scott Garren

## Loading and preprocessing the data

```{r}
setwd("~/!RR/project 1/RepData_PeerAssessment1")  
activity_data<-read.csv("activity.csv", colClasses=c("numeric", "Date", "character"))  
  
```  

## What is mean total number of steps taken per day?
 
```{r}
by_day<-split(activity_data, activity_data$date)
day_mean<-NULL
day_sum<-NULL
for (day in by_day) {  
  day_mean<-c(day_mean,mean(day$steps,na.rm=TRUE))  
  day_sum<-c(day_sum,sum(day$steps,na.rm=TRUE)) 
  }  
cat(" The means and median are different after imputing the missing data. The mean and median are the same which is quite a coincidence. ")
cat("mean of the daily steps=", mean(day_sum, na.rm=TRUE))  
cat("median of the daily steps=", median(day_sum, na.rm=TRUE))  
hist(day_sum, xlab="daily total steps", main="Histogram of daily step totals")  
```
  
## What is the average daily activity pattern?
```{r}
interval_ave<- function(act_data) {
  by_interval<-split(act_data, act_data$interval)  
  interval_mean<-NULL  

  for (interval in by_interval) {  
    interval_mean<-c(interval_mean,mean(interval$steps,na.rm=TRUE))  
    }
  return (interval_mean)
  }
int_ave<-interval_ave(activity_data)
plot(int_ave, type="l",xlab="average steps", main="Average of steps by interval")  

max_steps<-max(int_ave)
max_int<-activity_data[match(max_steps,int_ave),3]
cat("Maximum average steps was", max_steps, " in interval", max_int)

```

## Imputing missing values

I will replace all missing steps variables with the mean for that interval across the whole data set.

```{r}
for(i in 1:nrow(activity_data)) {  
    if (is.na(activity_data$steps[i])) {   
      activity_data$steps[i]<-interval_mean[(i-1)%%288+1]}   
    }   
    
by_day<-split(activity_data, activity_data$date)   
day_mean<-NULL   
day_sum<-NULL   
for (day in by_day) {  
  day_mean<-c(day_mean,mean(day$steps,na.rm=TRUE))  
  day_sum<-c(day_sum,sum(day$steps,na.rm=TRUE)) 
  }  
cat("mean of the daily steps=", mean(day_sum, na.rm=TRUE))  
cat("median of the daily steps=", median(day_sum, na.rm=TRUE))  
hist(day_sum, xlab="daily total steps", main="Histogram of daily imputed step totals")  
```   

## Are there differences in activity patterns between weekdays and weekends?
 Add a weekday factor field - TRUE>weekday, FALSE>weekend
 
```{r}
for(i in 1:nrow(activity_data)) {  
    if (as.numeric(strftime(activity_data$date[i],format="%u"))>5) { activity_data$weekday[i] <- FALSE
      } else activity_data$weekday[i] <- TRUE
      }
```

calculate means for weekdays and weekends

```{r}
weekday_interval_average<-interval_ave(activity_data[activity_data$weekday,])
weekend_interval_average<-interval_ave(activity_data[!activity_data$weekday,])
```

set up two row plot 

```{r}
par(mfcol=c(2,1))
plot(weekend_interval_average, type="l",xlab="time interval",ylab="average steps", main="Weekend average of steps by interval") 
plot(weekday_interval_average, type="l",xlab="time interval",ylab="average steps", main="Weekdays average of steps by interval")   

```